\thetitle

### 2021-04-22 writing/reviews/icfp21-rebuttal.md

The first part of our response addresses one selected theme from each review;
the second part addresses all remaining comments and questions individually.

What the experimental results say about the theory vs the practice of blame
================================================================================

Theoretical work that compares different checking and blame strategies for
gradual types (Greenman et al. ICFP 2018 and OOPSLA 2019) declares a clear
winner: Natural. Natural is the only semantics that is strongly type sound,
a complete monitor and, most importantly in the context of this paper, has
sound and complete blame assignment. The latter means that Natural blames
all and only those components that are relevant to a type impedance. In
contrast Transient may fail to blame some relevant components and may
blame irrelevant ones.

At first glance, the experiment of this paper seems to validate the
theoretical conclusions:

1. Natural's blame assignment is as good as the theory predicts; 
Natural blame helps locate impedance mismatches in almost all 
debugging scenarios. The only exceptions are (i) due to bugs in Typed
Racket and (ii) debugging scenarios that result in runtime errors with
unhelpful stacktraces from the underlying checks of Racket rather than 
Typed Racket's checks.

2. In general, Natural blame is better that Transient blame and both are
better than Erasure's stacktraces.

However, a closer inspection of the empirical results reveals that 
they paint a picture with gray zones that the black and white theoretical 
results are  not refined enough to capture:

1. There is a small but substantial number of debugging scenarios where 
Natural blame is not helpful but Transient blame is. These scenarios are
not all due to bugs in Typed Racket. Hence, the theory that deems natural
blame superior to that of Natural cannot explain them away.

2. While the theory claims that language designers should select Natural and
   its blames for maximum benefits, the results of the experiment indicate
   that Transient and Erasure are good enough for locating impedance
   mismatches in most cases. Hence, given the cost of Natural, the theory's answer
   fails to provide good enough guidance to language designers.

In sum the experiment in this paper reveals the mismatch between the theory and the practice
of blame. The latter refers to blame's actual utility in real
programs that run in a real implementation rather than artificial
examples in an idealized model.

As a result, there are two intertwined directions forward:

1. Revisiting the theory of blame to match the practice of blame.

2. Analyzing further the practice of blame beyond the scenarios this paper
   considers.

@@

Section 11 points to another space of scenarios that deserves
analysis: scenarios where mismatches arise due to mistakes in type
annotations rather than (or as well as) code.  Indeed, review B
astutely points out two more spaces worth exploring:

- programs where some components _cannot be typed_
- non-deterministic programs, which inject ephemeral bugs.

What is the origin of the rational programmer
================================================================================

Lazarek et al. (POPL'20) provide the kernel insight that one can simulate
an idealized programmer on a large corpus of programs to test the behavior
of a language feature. This paper hones this kernel into an experimental
framework that allows the isolation of confounding factors and,
furthermore, the comparison between different semantics for a feature in
the form of different but comparable experimental modes. The paper
christens the experimental framework the Rational Programmer.

@@

(if (integer? x) (use-int x) (use-bool x))
```

> How do you arrive at the number 72192 in Section 7? It is not equal to 756 times 96.
Indeed, 756 is a typo: it should be 752.

Review #45B
===========================================================================

>  - Which blame assignment strategy is "good" and which is not "good"?
TODO

>  - At first glance, the conclusion in lines 55-56 (starting with "Second, ...")
>    seems to contradict that in line 99-100 (starting with "neither is ...").  It would
>    be nicer to address them in a clearer manner.
Thanks for the feedback, we will rephrase the conclusion.

> - The paper often mixes gradual typing and migratory typing.  It is unclear how
>   it distinguishes them.
TODO

> - It is not fully explained how the given mutators change programs.  For
>   example, I cannot completely predict changes by the mutators deletion and
>   class:super.
In the specific case of `deletion`, the other most common place that
>   this mutator applies are so-called "implicit begins": places where
>   sequences are allowed implicitly by enclosing forms like `define`
>   or `cond`. 
So here are two more examples of `deletion` mutations:
```
(define (do-something) (step-1!) (step-2!))
~>
(define (do-something) (step-1!))

### 2021-04-22 writing/reviews/icfp21-rebuttal.md

Lazarek et al. (POPL'20) provide the kernel insight that one can simulate
an idealized programmer on a large corpus of programs to test the behavior
of a language feature. This paper hones this kernel into an experimental
framework that allows the isolation of confounding factors and,
furthermore, the comparison between different semantics for a feature in
the form of different but comparable experimental modes. The paper
christens the experimental framework the Rational Programmer.
=======

Replies to individual points
==================================================================

Review #45A
===========================================================================

> p5. one thing I wonder about is how this all hinges on the rational programmer not only taking the right action, but also taking it right? Is that exactly what the rational programmer idea presupposes? (I guess this reflects on your remark in Sec. 11 that one problem is with errors in the ascribed types themselves.).

Yes, you hit the nail on the head. 
One of the central parts of the rational programmer is that it fixes the decision of how to respond to a given scenario; 
so evaluating different ways of taking action boils down to defining and testing the corresponding rational programmers.

> When the rational programmer "fixes" a module and reruns, the implicit assumption seems to be that he runs the program in the exact same way. That would make the most sense to me, but you do not seem to come out and say so.

Yes you're right. Thanks, we will clarify this in the prose.

> It would help if you could explain how a mutator like negate-cond actually leads to a program
> that has more type-level mistakes than before. It does not seem to me this part of the code
> itself will now fail, but that changing the conditions makes certain previously excluded paths in the code feasible.

Thanks for pointing this out, we will expand the explanation and examples of mutators in the final paper.
In the specific cases of `negate-cond` and `force-cond`, these mutators lead to type-level mistakes when the program uses occurrence typing. 
For example, they could both lead to a mistake in a program like this:
```
(: use-int  (-> Integer Void))
(: use-bool (-> Boolean Void))
(: x        (U Integer Boolean))

@@

> How do you arrive at the number 72192 in Section 7? It is not equal to 756 times 96.

Indeed, 756 is a typo: it should be 752.

Review #45B
===========================================================================

>  - Which blame assignment strategy is "good" and which is not "good"?

TODO - Christos will do it.

>  - At first glance, the conclusion in lines 55-56 (starting with "Second, ...")
>    seems to contradict that in line 99-100 (starting with "neither is ...").  It would
>    be nicer to address them in a clearer manner.

Thanks for the feedback, we will rephrase the conclusion.

> - The paper often mixes gradual typing and migratory typing.  It is unclear how
>   it distinguishes them.

The paper uses the two terms interchangeably after the introduction, because the distinction between the two does not change the evaluation method.
Specifically, in the gradual typing setting the components are smaller than whole modules, and there are more possible intermediate types (incorporating Dyn) between untyped and fully-typed, both of which just end up making the configuration lattice larger.

> - In principle, Natural blame should always point out the faulty components
>   by the Wadler-Findler slogan.  However, the experiment shows it is not the
>   case.  Why?  Is it the same reason as in Lazarek et al.?  (Perhaps it is due
>   to a gap between theory and practice, but exposing a reason is crucial.)

(Combining this question with the related one from below next.)

We agree that the paper needs a bit more explanation to clarify how Natural blame can have failing
scenarios, we will update it in the final paper.  In short, Natural blame
trails can by stymied by unhelpful run-time errors.

Because debugging scenarios are only partially typed, some scenarios
produce an error from the runtime (e.g. `+` receiving a non-number)
before any of Natural's run-time type checks can discover a problem.
Run-time errors do not carry blame, only stacktraces, so the only way
to proceed (regardless of mode) is by using the stack information.  If
the stack proves unhelpful, then the trail fails.

> - It is not fully explained how the given mutators change programs.  For
>   example, I cannot completely predict changes by the mutators deletion and
>   class:super.

Thanks for the comment, we will expand the explanation and examples of mutators in the final paper.
In the specific case of `deletion`, the most common place that this mutator applies
(beyond the literal `begin` expressions the example illustrates)
are so-called "implicit begins": places where
sequences are allowed implicitly by enclosing forms like `define`
or `cond`.
Here are two more examples of `deletion` mutations for these situations:
```
(define (do-something) (step-1!) (step-2!))
~>
(define (do-something) (step-1!))

### 2021-04-22 writing/reviews/icfp21-rebuttal.md

The first part of our response addresses one major comment from each
reviewer.

Reviewer A says:
> the paper talks in the introduction of its top-level innovation being the
> idea of the rational programmer. But then they go on to say that the idea
> is actually taken from Lazarek. Which is it then?

Lazarek et al. (POPL'20) provide the kernel insight that one can simulate
an idealized programmer on a large corpus of programs to test the behavior
of a language feature. This paper hones this kernel into an experimental
framework. With the framework we can now:
1. isolate confounding factors. 
2. compare different semantics for gradual type boundaries in the form of 
   experimental modes.  
The paper christens this experimental framework the Rational Programmer.

@@

Lazarek et al. (POPL'20) provide the kernel insight that one can simulate
an idealized programmer on a large corpus of programs to test the behavior
of a language feature. This paper hones this kernel into an experimental
framework. With the framework we can now:
1. isolate confounding factors. 
2. compare different semantics for gradual type boundaries in the form of 
   experimental modes.  
The paper christens this experimental framework the Rational Programmer.

What the experimental results say about the theory vs the practice of blame
================================================================================

Reviewer B says:
> The conclusions of the second contribution described in the introduction
>  (around lines 41-103) are confusing for me.

Theoretical work that compares different checking and blame strategies for
gradual types (Greenman et al. ICFP 2018 and OOPSLA 2019) declares a clear
winner: Natural. Natural is the only semantics that is strongly type
sound, a complete monitor and, most importantly in the context of this
paper, Natural blames all and only those components that are relevant to a
type impedance mismatch (sound and complete blame). In contrast Transient
may fail to blame some relevant components and may blame irrelevant ones.
Hence, theory predicts that Natural is superior to Transient and Transient
is superior to Erasure.

At first glance, our experiments seem to validate these predictions:

1. When Natural issues blame it is helpful in almost all debugging
   scenarios. The problem is that many debugging scenarios result in
   runtime errors with unhelpful stacktraces from the underlying checks of
   Racket rather than Typed Racket boundary checks.

2. Natural blame looks better that Transient blame and both look better
   than Erasure's stacktraces.

However, a close inspection of the empirical data reveals that 
 a gray-shaded picture not the black and white print of theory:

1. There are some debugging scenarios where Natural blame is not helpful
   but Transient blame is. The theory cannot explain these scenarios.

2. While the theory claims that language designers should select Natural
   and its blame for maximum benefits, the data indicates that both
   Transient blame and Erasure stacktrace often suffice. Hence, given the
   cost of Natural (Takikawa et al. POPL 2016), the theory's suggestion
   has shown to be of questionable value.

In sum our experiments reveal the mismatch between the theory and the
practice of blame. The latter refers to blame's actual utility in real
programs that run in a real implementation rather than artificial examples
in an idealized model.

As a result, there are two intertwined directions going forward:

@@

1. Revisiting the theory of blame to match the practice of blame.

2. Empirically analyzing the practice of blame beyond the scenarios this
   paper considers. Section 11 points to another space of scenarios that
   deserves analysis: scenarios where mismatches arise due to mistakes in
   type annotations rather than (or as well as) code.  Indeed, review B
   astutely points out two more directions worth exploring in this
   space:
    - programs where some components _cannot be typed_;
    - non-deterministic programs.

@@

> When the rational programmer "fixes" a module and reruns, the implicit assumption seems to be that he runs the program in the exact same way. That would make the most sense to me, but you do not seem to come out and say so.

Yes you're right. Thanks, we will clarify this in the prose.

> It would help if you could explain how a mutator like negate-cond actually leads to a program
> that has more type-level mistakes than before. It does not seem to me this part of the code
> itself will now fail, but that changing the conditions makes certain previously excluded paths in the code feasible.

Thanks for pointing this out, we will expand the explanation and examples
of mutators in the final paper.  In the specific cases of `negate-cond`
and `force-cond`, these mutators lead to type-level mistakes when the
program uses occurrence typing.  For example, they could both lead to a
mistake in a program like this:

```
(: use-int  (-> Integer Void))
(: use-bool (-> Boolean Void))
(: x        (U Integer Boolean))

@@

Indeed, 756 is a typo: it should be 752.

Review #45B
===========================================================================

>  - Which blame assignment strategy is "good" and which is not "good"?

In this context, good means Natural and Transient. We will rephrase.

@@

>  - At first glance, the conclusion in lines 55-56 (starting with "Second, ...")
>    seems to contradict that in line 99-100 (starting with "neither is ...").  It would
>    be nicer to address them in a clearer manner.

Thanks for the feedback, we will rephrase the conclusion.

> - The paper often mixes gradual typing and migratory typing.  It is unclear how
>   it distinguishes them.

The paper uses the two terms interchangeably after the introduction,
because the distinction between the two does not change the evaluation
method.  Specifically, in the gradual typing setting the components are
smaller than whole modules, and there are more possible intermediate types
(incorporating Dyn) between untyped and fully-typed, both of which just
end up making the configuration lattice much much larger.

@@

> - In principle, Natural blame should always point out the faulty components
>   by the Wadler-Findler slogan.  However, the experiment shows it is not the
>   case.  Why?  Is it the same reason as in Lazarek et al.?  (Perhaps it is due
>   to a gap between theory and practice, but exposing a reason is crucial.)

(Combining this question with the related one from below next.)

> - Are failing Natural blame trails produced even for programs with impedance
>   mismatch?  (This question is related to the above issue with Natural blame.)

We agree that the paper needs a bit more explanation to clarify how
Natural blame can have failing scenarios, we will update it in the final
paper.  In short, Natural blame trails can by stymied by unhelpful
run-time errors from Racket. Because debugging scenarios are only
partially typed, some scenarios produce an error from the runtime (e.g.
`+` receiving a non-number) before any of Natural's run-time type checks
can discover a problem.  Runtime errors do not carry blame, only
stacktraces.

@@

> - It is not fully explained how the given mutators change programs.  For
>   example, I cannot completely predict changes by the mutators deletion and
>   class:super.

Thanks for the comment, we will expand the explanation and examples of
mutators in the final paper.  In the specific case of `deletion`, the most
common place that this mutator applies (beyond the literal `begin`
expressions the example illustrates) are so-called "implicit begins":
places where sequences are allowed implicitly by enclosing forms like
`define` or `cond`.  Here are two more examples of `deletion` mutations
for these situations:

```
(define (do-something) (step-1!) (step-2!))
~>
(define (do-something) (step-1!))

@@

(cond [(do-something?) (step-1!) (step-2!)]
      [else (void)])
~>
(cond [(do-something?) (step-1!)]
      [else (void)])
```

In the case of `class:super`, the mutator simply removes calls to
`super-new`, causing the class to no longer instantiate the superclass.

@@

> - The paper says that the experiment uses 72,192 sampled scenarios (line 802).
>   However, the total number of mutants (i.e., configuration lattices) is 756
>   and, from each lattice, 96 scenarios are sampled.  Thus, I guess 756*96 =
>   72576 scenarios are used.  Why does this mismatch happen?

Indeed, 756 is a typo: it should be 752.

> - The experiment samples multiple scenarios from a configuration lattice (line
>   800).  Does this mean the debugging process may start with a mix-typed program
>   and may not with a fully untyped one?  If so, why does the paper take such an
>   approach?

The debugging process can start with the fully untyped program but it
doesn't have to. The lattice contains the completely untyped program (bottom
configuration) and we sample from all the configurations where
the buggy module is untyped (including the fully untyped one).

@@

> - The paper often says that a blame system is (un)sound, but it is difficult for
>   me to identify what it precisely means.

The term refers to blame soundness from Greenman et al. (OOPSLA'19), but
we agree that for clarity the paper should define it in addition to the
reference.  Intuitively, a blame system is sound if it can only blame
components that have had control of the value witnessing a violation.  In
other words, the system cannot blame arbitrary components unrelated to the
witness of a contract violation.

@@

> - The paragraph starting at line 1122 is quite difficult to understand for me.
>   [...]
>   - What does the paper mean by "incomplete population of the blame map"?
>   - To understand the idea on the improvements of Transit, more explanations
>     on the usage of blame maps in Transient would be needed.

 We will improve the prose in this paper to clarify. Thank you for
 pointing this out.  The paper is referring here to the incomplete
 population of the blame map described by Greenman et al. ([11]:
 OOPSLA'19).

 For the details of how we have implemented the blame map and its
 population, there is a parallel experience report submission by the same
 set of authors describing the implementation of Transient in depth.

@@

>   - What "sophisticated typing features" are considered (line 589)?  It would be
>     crucial to confirm whether the proposed mutators are enough.

The features we have in mind appear in line 641. We will add a breakdown
of features per benchmark to the analysis of section 6.3. Our mutators do
target all of these.

@@

>   - The benchmarks are selected (line 647), but why?  The GPT benchmark suite of
>     Racket provides more examples.

The benchmarks in figure 5 are the ten with the largest dependency graphs
of the GTP suite. We selected those because small benchmarks have
short blame trails. We will rephrase L646-648 to make this clearer.

@@

> - Perhaps it is valuable for followers to share the experience on developing
>   mutators that are not interesting, .e.g., in the supplementary material.

We agree and would be happy to do so.

> - Why doesn't the paper adopt the author-year citation style?

This was a mistake typesetting the paper, we have fixed it. It didn't
affect the length of the paper

@@

> - L557: "if it does" --> "if it does not"?

You're right, thanks, we will fix it.

> - L685 "e.g., changing '*' to '/'":  Is this an example of the mutator that
>   "does not reliably lead to type errors"?

This is a refinement of the arithmetic mutator that does lead to type
errors, because given two integers `*` always produces an integer, but `/`
may produce a rational, which have different types in Typed Racket's
sophisticated numeric tower (St-Amour et al. PADL 2011).  We will clarify
this comment in the footnote.

@@

> - What is the total number of failing trails in Transient?

3594 out of 72192 trails fail for transient last blame, and 3591 for
first.

> - L1099 "as behavioral economics has shown more recently":  Is there a reference
>   to be cited?

The best reference we might use here is Kahnemann's popular book entitled
"Thinking Fast and Slow", which explains the problem well.  Notably with
respect to this work, the book criticizes classical (mathematical)
economic theory for assuming that people act rationally all of the time,
which demonstrates how our work is complementary to human studies and
theory. All are necessary.

### 2021-04-23 writing/reviews/icfp21-rebuttal.md

The first part of our response addresses one major comment from each
reviewer.

What is the origin of the rational programmer
================================================================================

Reviewer A says:
> the paper talks in the introduction of its top-level innovation being the
> idea of the rational programmer. But then they go on to say that the idea
> is actually taken from Lazarek. Which is it then?

Lazarek et al. (POPL'20) provide the kernel insight that one can simulate
an idealized programmer on a large corpus of programs to test the behavior
of a language feature. This paper hones this kernel into an experimental
framework. With the framework we can now:

1. isolate confounding factors;
2. compare different semantics for gradual type boundaries in the form of 
   experimental modes.

The paper christens this experimental framework the Rational Programmer.
We have already revised the prose accordingly.

@@

What the experimental results say about the theory vs the practice of blame
================================================================================

Reviewer B says:
> The conclusions of the second contribution described in the introduction
>  (around lines 41-103) are confusing for me.

Theoretical work that compares different checking and blame strategies for
gradual types (Greenman et al. ICFP 2018 and OOPSLA 2019) declares a clear
winner: Natural.

- _Natural_ is the only semantics that is type sound in the conventional sense,
  completely monitors boundaries, and, most importantly here, blames all and
  only those components that are relevant to a type impedance mismatch (sound
  and complete blame).

- By contrast, _Transient_ may fail to blame some relevant components and may
  blame irrelevant ones.

- Erasure doesn't even assign checks to type boundaries.

Hence, theory predicts that Natural is superior to Transient and Transient is
superior to Erasure.

Our experiments only appear to validate these predictions:

1. When Natural issues blame it is helpful in almost all debugging scenarios.

@@

2. Natural blame looks better that Transient blame and both look better than
   Erasure's stacktraces.

However, a close inspection of the empirical data reveals that 
 a gray-shaded picture not the black and white print of theory:

1. One problem is that many debugging scenarios result in runtime errors with
   unhelpful stacktraces from the underlying checks of Racket rather than Typed
   Racket boundary checks.

1. Additionally, for some debugging scenarios Natural blame is not helpful but
   Transient blame is. The theory cannot explain these scenarios at this point.

2. Finally, while the theory claims that language designers should select
   Natural and its blame for maximum benefits, the data indicates that both
   Transient blame information and Erasure stacktraces often suffice.  Given the
   large cost of Natural (Takikawa et al. POPL 2016), theory has has shown
   itself to be of questionable value -- at this time.

In sum our experiments reveal a mismatch between theory and practice where blame
may or may not matter for real programs running on real implementations. This is
in contrast to theoretical papers, which symbol-push small terms thru a reduction 
semantics in a tiny model of PLs (even if these models exhibit all essential 
attributes).

As a result, we expect to invest in two intertwined research efforts going forward:

@@

1. Revisiting the theory of blame to match the practice of blame.

2. Empirically analyzing the practice of blame beyond the scenarios this
   paper considers. Section 11 points to another space of scenarios that
   deserves analysis: scenarios where mismatches arise due to mistakes in
   type annotations rather than (or as well as) code.

   Indeed, the review astutely points out two more directions worth exploring:

    - programs where some components _cannot be typed_;
    - non-deterministic programs.

### 2021-04-23 writing/reviews/icfp21-rebuttal.md

The first part of our response addresses one major comment from each review.

The origin of the rational programmer
================================================================================

Review A says:
> the paper talks in the introduction of its top-level innovation being the
> idea of the rational programmer. But then they go on to say that the idea
> is actually taken from Lazarek. Which is it then?

@@

Lazarek et al. (POPL'20) provide the kernel insight that one can simulate
an idealized programmer on a large corpus of programs to test the behavior
of a language feature. This paper hones this kernel into an experimental
framework. With the framework we can now:

1. isolate confounding factors;
2. compare different semantics for gradual type boundaries in the form of
   experimental modes.

The paper christens this experimental framework the Rational Programmer.
We have already revised the prose accordingly.

What the results say about the theory vs practice of blame
================================================================================

Review B says:
> The conclusions of the second contribution described in the introduction
>  (around lines 41-103) are confusing for me.

@@

Our experiments only appear to validate these predictions:

1. When Natural issues blame it is helpful in almost all debugging scenarios.

2. Natural blame looks better that Transient blame and both look better than
   Erasure's stacktraces.

However, a close inspection of the empirical data reveals a gray-shaded picture
instead of the black and white print of theory:

@@

1. One problem is that many debugging scenarios result in runtime errors with
   unhelpful stacktraces from the underlying checks of Racket rather than Typed
   Racket boundary checks.

1. Additionally, for some debugging scenarios Natural blame is not helpful but
   Transient blame is. The theory cannot explain these scenarios at this point.

2. Finally, while the theory claims that language designers should select
   Natural and its blame for maximum benefits, the data indicates that both
   Transient blame information and Erasure stacktraces often suffice.  Given the
   large cost of Natural (Takikawa et al. POPL 2016), theory has has shown
   itself to be of questionable value -- at this time.

In sum, our experiments reveal a mismatch between theory and practice where blame
may or may not matter for real programs running on real implementations. This is
in contrast to theoretical papers, which symbol-push small terms through a reduction
semantics in a tiny model of PLs (even if these models exhibit all essential
attributes).

@@

    - programs where some components _cannot be typed_;
    - non-deterministic programs.

Replies to individual points
==================================================================

Review #45A
===========================================================================

> The table in section 2 looks rather ugly

We will find another way to present the information.

@@

Review #45B
===========================================================================

>  - Which blame assignment strategy is "good" and which is not "good"?

In this context, good means Natural and Transient. We will rephrase.

>  - Why does the paper conclude that "the existing theory does not predict
>    practice properly"?  Figure 8 shows that the theory (Natural) works well.
>  - What does the paper mean by "The existing practice may need additional
>    experiments"?  I have not found any evidence that confirms this claim in the
>    paper.
>  - What does "practice" means here?

We hope the start of the response addresses these questions adequately.

@@

>  - At first glance, the conclusion in lines 55-56 (starting with "Second, ...")
>    seems to contradict that in line 99-100 (starting with "neither is ...").  It would
>    be nicer to address them in a clearer manner.

Thanks for the feedback, we will rephrase the conclusion.

> - The paper often mixes gradual typing and migratory typing.  It is unclear how
>   it distinguishes them.

The paper uses the two terms interchangeably after the introduction,
because the distinction between the two does not change the evaluation
method.  Specifically, in the gradual typing setting the components are
smaller than whole modules, and there are more possible intermediate types
(incorporating Dyn) between untyped and fully-typed, both of which just
end up making the configuration lattice much larger.

@@

> - In principle, Natural blame should always point out the faulty components
>   by the Wadler-Findler slogan.  However, the experiment shows it is not the
>   case.  Why?  Is it the same reason as in Lazarek et al.?  (Perhaps it is due
>   to a gap between theory and practice, but exposing a reason is crucial.)

(Combining this question with the related one from below next.)

> - Are failing Natural blame trails produced even for programs with impedance
>   mismatch?  (This question is related to the above issue with Natural blame.)

We agree that the paper needs a bit more explanation to clarify how Natural
blame can have failing scenarios, we will update it in the final paper. In
short, Natural blame trails can by stymied by unhelpful errors from Racket's
runtime. Because debugging scenarios are only partially typed, some scenarios
produce an error from the runtime (e.g. `+` receiving a non-number) before any
of Natural's run-time type checks can discover a problem. Errors from the
runtime do not carry blame, only stacktraces.

@@

> - The paper says that the experiment uses 72,192 sampled scenarios (line 802).
>   However, the total number of mutants (i.e., configuration lattices) is 756
>   and, from each lattice, 96 scenarios are sampled.  Thus, I guess 756*96 =
>   72576 scenarios are used.  Why does this mismatch happen?

Indeed, 756 is a typo: it should be 752.

> - The experiment samples multiple scenarios from a configuration lattice (line
>   800).  Does this mean the debugging process may start with a mix-typed program
>   and may not with a fully untyped one?  If so, why does the paper take such an
>   approach?

The debugging process can start with the fully untyped program but it
doesn't have to. The lattice contains the completely untyped program (bottom
configuration) and we sample trail roots from all the configurations where
the buggy module is untyped (including the fully untyped one).

@@

We will improve the prose in this paper to clarify. Thank you for pointing this
out. The paper is referring here to the incomplete population of the blame map
described by Greenman et al. (OOPSLA'19).

For the details of how we have implemented the blame map and its population,
there is a parallel experience report submission by the same set of authors
describing the implementation of Transient in depth.

>   - What "sophisticated typing features" are considered (line 589)?  It would be
>     crucial to confirm whether the proposed mutators are enough.

The features we have in mind appear in line 641. We will add a breakdown of
features per benchmark to the analysis of section 6.3 to clarify that our
mutators target all of these.

@@

>   - The benchmarks are selected (line 647), but why?  The GPT benchmark suite of
>     Racket provides more examples.

The benchmarks in figure 5 are the ten with the largest dependency graphs
of the GTP suite. We selected those because small benchmarks have
short blame trails. We will rephrase L646-648 to make this clearer.

> - It seems that the experiment implicitly supposes a few assumptions on
>   benchmarks.  First, the original benchmarks must be fully typed.  Thus more
>   experiments on mix-typed programs where some components cannot be typed may be
>   needed.  Second, the benchmarks are supposed to be deterministic, which I find
>   from the definitions of trails.  If it is the case, while I do not think these
>   restrictions have to be lifted in the paper, it would be nice to expose them.

You are absolutely right, we will clarify these assumptions about the
benchmarks in the paper. They are inherited from the other
publications on the gradual benchmark suite.

### 2021-04-23 writing/reviews/icfp21-rebuttal.md

1. One problem is that many debugging scenarios result in runtime errors with
   unhelpful stacktraces from the underlying checks of Racket rather than Typed
   Racket boundary checks.

1. Additionally, for some debugging scenarios Natural blame is not helpful but
   Transient blame is. The theory cannot explain these scenarios at this point.

2. Finally, while the theory claims that language designers should select
   Natural and its blame for maximum benefits, the data indicates that both
   Transient blame information and Erasure stacktraces often suffice.  Given the
   large cost of Natural (Takikawa et al. POPL 2016), theory has has shown
   itself to be of questionable value -- at this time.

In sum, our experiments reveal a mismatch between the theory and the
practice of blame, where practice refers to blame's actual utility in real
programs that run in a real implementation rather than artificial examples
in an idealized model.

### 2021-04-23 writing/reviews/icfp21-rebuttal.md

The first part of our response addresses one major comment from each review.

@@

What is the origin of the rational programmer
================================================================================

Review A says:
> the paper talks in the introduction of its top-level innovation being the
> idea of the rational programmer. But then they go on to say that the idea
> is actually taken from Lazarek. Which is it then?

@@

1. isolate confounding factors;
2. compare different semantics for gradual type boundaries in the form of 
   experimental modes.

The paper christens this experimental framework the Rational Programmer.
We have already revised the prose accordingly.

What the experimental results say about the theory vs the practice of blame
================================================================================

Review B says:
> The conclusions of the second contribution described in the introduction
>  (around lines 41-103) are confusing for me.

### 2021-04-24 writing/reviews/icfp21-rebuttal.md

This paragraph refers to a Python translation of the Racket `Sieve`
benchmark; it is confusingly worded, we will adjust the phrasing.

> - L118: "(at the mid-level on the left)[] imports" (the comma is removed)
> - L332 "soundness mechanisms":  What they mean is somewhat unclear.  Please
>   consider clarifying.
> - L347 "the rational programmer's effort": This was unclear at the first
>   reading.  It would be nice to briefly explain it here.
> - L451: "denotes the module [of P] that s blames"
> - L970: "the experiment provides [an] evidence that"
> - L1198 "But just because...":  It is difficult for me to find what this
>   sentence wants to say.  Please consider rephrasing.

Thanks for these comments, we will incorporate them to improve the prose.

Review #45C
===========================================================================

> My only concern at this point is that we need to believe that the authors have
> implemented the Transiet semantics correctly in Racket, though no details are
> offered. The paper refers to another submission, so perhaps some details can
> be found in there.

TODO

> I am wondering whether the approach that the paper calls Transient First could
> be regarded as Transient Early, since it points to an earlier part of the
> code. Similarly Transient Last could be Transient Late, but you would have to
> see whether this terminology really fits.

TODO

> The only thing that I am concerned about at this point is that, later in the
> paper, it is shown that "most cases the programmer need to type a single
> module to debug a scenario," which makes me wonder whether the authors did not
> generate difficult enough scenarios. If that is the case, what is the problem?

TODO

> My only issue at this point is that the reader is left with numnbers only,
> while I think a reader would like to see selected examples, for example about
> which blame labels have been provided by Natural and which by Transient, so to
> see that one approach took a longer path, as well as similar examples the
> reader can learn from. Numbers do not seem to teach the whole story in this
> part of the paper.

TODO

### 2021-04-24 writing/reviews/icfp21-rebuttal.md

The origin of the rational programmer
================================================================================

Review A says:
> the paper talks in the introduction of its top-level innovation being the
> idea of the rational programmer. But then they go on to say that the idea
> is actually taken from Lazarek. Which is it then?

Lazarek et al. (POPL'20) provide the kernel insight that one can simulate
an idealized programmer on a large corpus of programs to test the behavior
of a language feature. This paper hones this kernel into an experimental
framework. With the framework we can now:

1. isolate confounding factors, such as whether the utility of blame is due
   to blame tracking or just due the checks;
2. compare different semantics for gradual type boundaries in the form of
   experimental modes.

@@

2. Empirically analyzing the practice of blame beyond the scenarios this
   paper considers. Section 11 points to another space of scenarios that
   deserves analysis: scenarios where mismatches arise due to mistakes in
   type annotations rather than (or as well as) code.

   Indeed, the review astutely points out two more directions worth exploring:

    - programs where some components _cannot be typed_;
    - non-deterministic programs.

Are the debugging scenarios difficult?
================================================================================

Review C says:
> The only thing that I am concerned about at this point is that, later in the
> paper, it is shown that "most cases the programmer need to type a single
> module to debug a scenario," which makes me wonder whether the authors did not
> generate difficult enough scenarios. If that is the case, what is the problem?

The bugs that we inject are non-trivial. To provide some evidence, we sampled a
few (~24,000) scenarios from three middle-of-the-road benchmarks (acquire,
take5, kcfa) and the median stack has 7 program modules, with the sizes ranging
from 3 to 10. This indicates that these bugs involve significant interaction
among different parts of the program.

One could argue that the benchmark programs are too small, because they consist
of about 10-14 modules. This is a decent size but not a truly large one, so
perhaps even larger benchmark programs are called for, though we conjecture that
results will remain roughly the same because the programming style is
representative (it includes a good number of higher-order idioms, for both
functions and classes; it includes a fair number of call-back scenarios across
module boundaries and storage of objects to be called later -- after the storage
call returns -- but again not too many).

@@

Thanks for these comments, we will incorporate them to improve the prose.

Review #45C
===========================================================================

> My only concern at this point is that we need to believe that the authors have
> implemented the Transiet semantics correctly in Racket, though no details are
> offered. The paper refers to another submission, so perhaps some details can
> be found in there.

Our implementation of Transient is faithful to Reticulated Python; the parallel
experience report submission by the same set of authors describes the
implementation of Transient in depth.

@@

> I am wondering whether the approach that the paper calls Transient First could
> be regarded as Transient Early, since it points to an earlier part of the
> code. Similarly Transient Last could be Transient Late, but you would have to
> see whether this terminology really fits.

Thanks, we will consider this alternative name for the final paper.

@@

> My only issue at this point is that the reader is left with numnbers only,
> while I think a reader would like to see selected examples, for example about
> which blame labels have been provided by Natural and which by Transient, so to
> see that one approach took a longer path, as well as similar examples the
> reader can learn from. Numbers do not seem to teach the whole story in this
> part of the paper.

This is a good idea, thanks. In the final paper we will try to incorporate one
or more examples that illustrate the difference in trail between the three
mechanisms.

### 2021-04-24 writing/reviews/icfp21-rebuttal.md

Review #45A
===========================================================================

> The table in section 2 looks rather ugly

We will find another way to present the information.

> p5. one thing I wonder about is how this all hinges on the rational programmer
> not only taking the right action, but also taking it right? Is that exactly
> what the rational programmer idea presupposes? (I guess this reflects on your
> remark in Sec. 11 that one problem is with errors in the ascribed types
> themselves.).

@@

Yes, you hit the nail on the head.  One of the central parts of the
rational programmer is that it fixes the decision of how to respond to a
given scenario; so evaluating different ways of taking action boils down
to defining and testing the corresponding rational programmers.

> When the rational programmer "fixes" a module and reruns, the implicit
> assumption seems to be that he runs the program in the exact same way. That
> would make the most sense to me, but you do not seem to come out and say so.

@@

Yes you're right. Thanks, we will clarify this in the prose.

> It would help if you could explain how a mutator like negate-cond actually
> leads to a program that has more type-level mistakes than before. It does not
> seem to me this part of the code itself will now fail, but that changing the
> conditions makes certain previously excluded paths in the code feasible.

@@

> My only concern at this point is that we need to believe that the authors have
> implemented the Transiet semantics correctly in Racket, though no details are
> offered. The paper refers to another submission, so perhaps some details can
> be found in there.

Our implementation of Transient is faithful to Reticulated Python; the parallel
experience report submission by the same set of authors describes the
implementation of Transient in depth.

> I am wondering whether the approach that the paper calls Transient First could
> be regarded as Transient Early, since it points to an earlier part of the
> code. Similarly Transient Last could be Transient Late, but you would have to
> see whether this terminology really fits.

Thanks, we will consider these alternative names for the final paper.

@@

> My only issue at this point is that the reader is left with numnbers only,
> while I think a reader would like to see selected examples, for example about
> which blame labels have been provided by Natural and which by Transient, so to
> see that one approach took a longer path, as well as similar examples the
> reader can learn from. Numbers do not seem to teach the whole story in this
> part of the paper.

This is a good idea, thanks. In the final paper we will try to incorporate one
or more examples that illustrate the difference in trails between the three
mechanisms.
